---
title: "Floral development and pollination of _Centropogon granulosus_ (Campanulaceae) in an Andean cloud forest"
author:
date: ""
output: 
  word_document:
    fig_caption: yes
    reference_docx: centropogon_biology_reference.docx
  # html_document:
  #   number_sections: no
  #   theme: flatly
   # pdf_document:
   #  latex_engine: xelatex
bibliography: centropogon_biology.bib
csl: cell.csl
link-citations: true #'yes' for html, 'true' for pdfs
linkcolor: blue
urlcolor: blue
citecolor: blue


---

**Author Names**: <br> 
Mannfred M.A. Boehm$^{1,2,*}$, David Guevara Apaza$^{3}$ + others

$^*$**Corresponding Author**: mannfred.boehm@ubc.ca

**Author affiliations**: <br> 
$^1$Department of Botany, University of British Columbia,
3156 - 6270 University Boulevard, Vancouver, BC, Canada, V6T 1Z4. <br>

$^2$Biodiversity Research Centre, University of British Columbia, 2212 Main Mall, Vancouver, BC, Canada, V6T 1Z4. <br>

$^3$Facultad de Ciencias Biológicas, Universidad San Antonio Abad del Cusco, Av. de La Cultura 773, Cusco 08000, Peru. 

**Submission Type**: Note

**Abstract**: (200 max for Botany) 

**Main Body**: (4200 max for Botany) 

**Figures**: # Figures

**Keywords**: (3-6 for Botany)



\newpage


#### Acknowledgements

Servicio Nacional Forestal y de Fauna Silvestre (SERFOR) of Peru.
Herbarium faculty and staff at the Universidad Nacional de San Antonio Abad del Cusco (UNSAAC).
Staff of the Gallieto de las Rocas Lodge. 
B. Freeman, M.N. Scholer, and E. Llacta-Cuno 


\newpage

#### Introduction 

The Andes comprise the highest mountains in the western hemisphere and are one of the most biodiverse regions on Earth [@myers_2000; @hoorn_2018]. With a mean elevation of 4000 m, the Andes encompass a wide environmental gradient ranging from warm Amazonian rainforests, to dry, montane grasslands. During Andean orogeny, numerous plant clades were either uplifted [@heads_2019] or migrated and diversified into the newly formed niche space, e.g. _Lupinus_ L. [Fabaceae, @hughes_2006], centropogonids [Lobelioideae, @lagomarsino_2016], epiphytic orchids [Epidendroideae, @perez_2017]. Today, about 10% of global angiosperm diversity is found in the Andes eco-region [@joppa_2011]. 

&emsp;&emsp; Plant-pollinator co-evolution offers a powerful model of the rapid diversification of the Andean flora [e.g. @kay_2005; @lagomarsino_2017; @serrano_2017]. As a plant species expands into a novel montane environment, floral traits may drift, evolve via pleiotropy, or adapt to a pre-existing pollinator community [@harder_2009; @smith_2016]. Subsequently, divergence in floral phenotype can be reinforced when reciprocal adaptation between the plant and pollinator promotes floral isolation [sensu @grant_1949]. However, despite compelling phenotypic and phylogenetic evidence, many putative cases of pollinator-driven plant diversification lack direct observations of floral visitation. In this study, we aimed to test the theory of hummingbird-mediated diversification in the eucentropogonids [sensu @lagomarsino_2014] by recording pollinator visits, performing a pollinator exclusion experiment, and synthesizing evidence from related studies. 

&emsp;&emsp; Here, we focus on the Andean bellflower _Centropogon granulosus_ C.Presl and buff-tailed sicklebill, _Eutoxeres condamini_ Bourcier (Phaethornithinae). The Andean bellflowers (*Centropogon* C. Presl), along with *Siphocampylus* Pohl and *Burmeistera* H.Karst & Triana form a South American clade of Campanulaceae known as the ‘centropogonids’ [@lagomarsino_2014]. The centropogonids arose in the last 5 million years and have diversified into over 550 species. This rapid radiation has been attributed to increased speciation rates driven by the Andean uplift, simultaneous with decreased extinction rates during periods of global cooling [@lagomarsino_2016]. The Andean clade _Centropogon_ section _Centropogon_ (McVaugh 1949) or the ‘eucentropogonids’ (38 spp., Lagomarsino et al 2014) are distinct because of their elongated, hook-shaped flower tube, and the presence of a scale made of fused hairs that serves as a lever to deposit pollen. Both traits are likely adaptations to facilitate pollination by sicklebill hummingbirds [@stein_1987]. Generally, the elevational distribution of the eucentropogonids is mid- to high-montane (Figure 1). 

&emsp;&emsp; Sicklebill hummingbirds (*Eutoxeres* Reichenbach) constitute two species of large hermits (Phaethornithinae) that are defined by their dramatically curved bill. Their unique bill shape is likely an adaptation to specialize on those _Centropogon_ and _Heliconia_ with similarly curved flowers. The hermits diverged early in the evolutionary history of hummingbirds [~26 Ma, @mcguire_2014], and represent one of two major ecotypes of hummingbirds: the trapliners. There are currently 36 extant hermit species [@mcguire_2014].  _Eutoxeres_ diverged 21.45 Ma, and like other hermits, probably co-adapted to _Heliconia_ before major uplifts of the Andes [@abrahamczyk_2017]. At present, _Eutoxeres_ has the largest range of any hermit hummingbird&mdash;it can be found in its ancestral range, the Amazonian lowlands, as well as the foothills and mid-montane habitats of the Andes up to 2800 m [@schulenberg_2010]. 

&emsp;&emsp; Because _Heliconia_ and the hermit hummingbirds co-existed before the emergence of the eucentropogonids, it is likely that sicklebills and _Heliconia_ subgenus _Stenochlamys_ and/or _Heliconia_ subgenus _Griggsia_ co-evolved extreme bill and flower curvature in lowland Amazonia, before the uplift of the Andes [@stein_1987; @abrahamczyk_2017]. During a recent period of major Andean uplift (~5 Ma) _Centropogon_ colonized and radiated into the novel montane habitat, with some species adapting to sicklebill pollination. This meditated the elevational expansion of sicklebills, and explains why they are able to exist outside of the range of _Heliconia_. However, the physiological pre-adaptations that allowed sicklebills to expand, and not other hermits, remains unknown [@stiles_2004]. This scenario presents the possibility of a plant-hummingbird interaction cline, with _Centropogon_ and _Eutoxeres_ co-evolving in the absence of _Heliconia_ at medium to high elevations (Figure 1), and _vice versa_ in the lowlands. The distribution of putatively sicklebill-pollinated _Centropogon_ and _Heliconia_ suggests that above ~1500 m, there are few Heliconia-Sicklebill interactions (Figure 2). 

&emsp;&emsp; While trait-matching and anecdotal reports strongly suggest that sicklebills are pollinators of _Centropogon_, empirical evidence remains absent. While the pollination syndrome concept has proven useful in inferring pollinators from floral phenotypes [but see: @ollerton_2009], a complete theory of pollinator-mediated plant diversification ultimately requires _in situ_ observations of visitation, and some evidence of pollination. Moreover, without a dedicated focus on the organisms' natural history, it cannot be ruled out that yet-unknown floral visitors directly or indirectly interact with the pollinator. All said, we aim to provide some missing fundamental information on this fascinating and increasingly well-studied pollination system. 

\newpage

#### Materials and Methods

_Field Site_
&emsp;&emsp; We based our field work at the Gallieto de las Rocas Lodge, in the Kosñipata Valley, Peru (-13.055, -71.548 DD). Research Permit No. 0441-2017 was administered by the Servicio Nacional Forestal y de Fauna Silvestre (SERFOR). In this study we focused on the pollination biology of _Centropogon granulosus_ C.Presl. While other eucentropogonid species were present at the study site (Figure 1), _C. granulosus_ was the most widespread and accessible for study. We located and recorded the locations of _C. granulosus_ individuals in a preliminary survey [@boehm_2018].


_Pollinator Observations_
&emsp;&emsp; In this study, we deployed camera traps near five individual _Centropogon granulosus_ vines. If no floral visitors were recorded within three days, the camera traps were moved to different _Centropogon_ individuals, totaling n=???. Where floral visitors were recorded, we performed targeted (manual) video recording to better document visitation behaviour. 

_Pollinator Exclusion Experiment_
&emsp;&emsp; Following the methods and designs of Sun et al. [-@sun_2017], we constructed and deployed wire cages covering one inflorescence each from six _Centropogon_ individuals. Wire cages prevent hummingbirds from contacting the flowers while allowing invertebrates to move freely. An additional ten inflorenscences were monitored as controls. Each inflorescence produced on average 11 and 13 flowers, respectively. Using the control flowers, we defined five discrete stages of floral development (Table 1). These stages were used to quantify and compare the developmental trajectories of the control and pollinator-excluded flowers. Monitoring for a sample stopped when 1) berry development completed (stage E), 2) the flower died prematurely, or 3) the study period ended. Flowers that did not reach berry development were treated as right-censored [@allison_2014]. Daily observations were recorded between August 17 and September 20, 2017. Flower development data was inputted to R v3.6.3 [@R_2017] and organized using `here` v0.1 [@muller_2017] and `tidyverse` v1.3.0 [@wickham_2019] 

&emsp;&emsp; Flowering stage data was staggered because each flower developed independently. The data was also fragmented because some samples abscissed prematurely due to herbivory or weather. The flower development data was therefore a 'censored' dataset i.e. some subjects left the study before reaching maturity [@allison_2014]. Although manually aligning the stage data is possible, for convenience we used a multiple sequence alignment protocol to automate the process. To do this, we ran the stage data through ClustalW implemented in the R package `msa` v.3.9 [@bodenhofer_2015] with a neutral (identity) substitution matrix. Aligned developmental sequence data was then used to compute a consensus (mean) sequence for each treatment &mdash; this was done using the `ConsensusSequence()` function in `DECIPHER` v2.14.0 [@wright_2016]. 

&emsp;&emsp; We then conducted an event history analysis to compare the probability of post-anthesis survival between pollinator-excluded and untreated flowers. To do this, we counted the number of days elapsed from anthesis (Stage D) through berry development (Stage E). Daily survival probability (from anthesis) was calculated using the `survival` package v3.1 [@therneau_2015]. 


\newpage

#### Results: 

_Pollinator Exclusion and Floral Development_
We identified and described five discrete stages of floral development in _C. granulosus_ (Table 1). These were used to compare differences in development (especially fruit set) between control and pollinator-excluded flowers. For all samples, individuals that had hummingbirds excluded did not produce berries. 


_Pollinator Observations_



_Pollinator Exclusion Experiment_

In studying the development and pollination of _C. granulosus_ flowers, we noted some natural history traits yet published elsewhere. First, the blah blah.. 
  
  
  - Figure 2. Developmental stages of Centropogon. 
  - Figure 3. Pollination by Sicklebills
  - Discuss things that live in or around the flowers (ants (cite Stein and LauraLago), mites (cite MacGregor), arachnids)

```{r include=FALSE} 
knitr::opts_chunk$set(echo=FALSE, warning=FALSE, message=FALSE)

```



```{r table1}

library(googlesheets4)

sheets_auth(email='mannfredboehm@gmail.com')
table1 <- read_sheet('1VzmXOzNcNKxmyFNfu_fBt6eqI4MsqAt3XuXTVmFJhlo', col_names=TRUE)

knitr::kable(table1, caption = 'Table 1: Stages of floral development in _Centropogon granulosus_.')

```


# References
  

